The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-09-26 article.tex
Modifed some of the intro

Open Source cheminformatics has made significant steps forward recently~\cite{OBoyle2011b}.
The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
For example, new tools expose CDK functionality in larger platforms such as Cinfony~\cite{OBoyle2008},
rcdk~\cite{Guha2007}, and plugins for Taverna~\cite{Truszkowski2011},
KNIME~\cite{Beisken2013}, ChemViz2 for Cytoscape~\cite{ChemViz2}, and LICSS for
Microsoft Excel~\cite{Lawson2012}.
Furthermore, projects with specific functionality have emerged building on the CDK, including
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
Other tools used previous versions already, and follow
newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. While the above
tools suggest that the CDK is primarily used to create general tools, it is
important to note that it is also used to solve specific questions, like finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
defining fingerprints for organic reactions~\cite{Segler2016}.

However, previous version of the CDK were written by a community with specific applications
in mind, of which structure elucidation was one. It could therefore be rather competitive
in some cheminformatics domains, such as fingerprinting~\cite{Clark2014,Cannon2006}, in other parts
it could be inadequate, such in the capturing of stereochemistry.

>>>>

The Open Source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Adoption of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
defining fingerprints for organic reactions~\cite{Segler2016}.

While the CDK has purported to be a general purpose cheminformatics
toolkit, older versions were designed by a community with specific
applications, primary among them being structure elucidation. In
addition, an implicit goal of previous versions was to have the CDK
serve as an educational resource to enable students of cheminformatics
to understand the underlying algorithms. This resulted in certain
functionalities, such as fingerprinting~\cite{Clark2014,Cannon2006},
receiving more attention than others, such as stereochemistry. The
outcome was significant variance in performance and features
throughout the toolkit.
